---
title: "Assignment 1"
description: |
  Biostats624 assignment 1
author:
  - name: Johnathan He
    affiliation: Epidemiology, Johns Hopkins
date: "`r Sys.Date()`"
output: distill::distill_article
---


```{r "setup", include = FALSE}
require("knitr")
opts_chunk$set(warning = FALSE, message = FALSE, echo = TRUE)
opts_knit$set(root.dir = "D:/OneDrive - Johns Hopkins/Course/140.624.71 - Statistical Methods in Public Health IV/assignments/biostat624-assignment/Assignment 1")
```


```{r}
require(tidyverse)

setwd("D:/OneDrive - Johns Hopkins/Course/140.624.71 - Statistical Methods in Public Health IV/assignments/biostat624-assignment/Assignment 1")
```


# Section 1


```{r}
wm <- read_csv("./DATA/words_matter_assg1_data.csv")
```


1.	How many observations are in the dataset?
2.	How many variables are in the dataset?
3.	What type of variable is ppage?  


### Taking stock of what is in the data set: the `str()` command (str = “structure”)


```{r}
str(wm)
```


4.	How many unique values of ppage are there?


```{r}
length(unique(wm$ppage))
```


### Assessing missing values: `is.na` function


5.	How many observations are missing a value for ppage?


```{r}
count(wm, is.na(ppage))
table(is.na(wm$ppage))
sum(is.na(wm$ppage))
```


6.	What type of variable is ppgender?  
7.	How many unique values of ppgender are there?
8.	How many observations are missing a value for ppgender?
9.	Based  on the results from these commands can you ascertain how many males are in the data sample?


```{r}
class(wm$ppgender)
str(wm$ppgender)
table(is.na(wm$ppgender))
length(unique(wm$ppgender))	

count(wm, ppgender)
```


### Labelling Variable Names with Descriptive Text


```{r}
library(expss)  	
wm <- apply_labels(wm,
                   caseid = "respondent id",
                   group = "Experimental Group",
                   ppeducat = "Highest Education Level (Categorical)",
                   ppethm = "Race and Ethnicity",
                   ppgender = "Biological Sex",
                   ppage = "Age (Years)")
```


```{r}
str(wm)
```


### Labeling Values of Numeric Variables


10.	 How do the values ppgender display – as numbers or words?   
11.	   What does the [1:15] addition to wm$ppgender do?


```{r}
wm <- wm %>%
    mutate(ppgender = factor(ppgender,
                             levels = c(1, 2),
                             labels = c("Male", "Female")))

table(wm$ppgender)
count(wm, ppgender)
```


```{r}
wm$ppgender[1:15]
```


12.	What does the addition of `as.numeric` to `wm$ppgender[1:15]` do?


```{r}
as.numeric(wm$ppgender[1:15])
```


### Now for some preliminary analyses


13.	What percentage of the sample are males?
14.	How many respondents are in education level category 2?  


```{r}
table(wm$ppgender)
prop.table(table(wm$ppgender))
```


```{r}
count(wm, ppgender)
#prop.table(count(wm, ppgender))
```


```{r}
table(wm$ppeducat)	
prop.table(table(wm$ppeducat))
```


```{r}
count(wm, ppeducat)
prop.table(count(wm, ppeducat))
```


15.	What is mean age of the subjects in the sample?
16.	The youngest person(s) in the sample is/are how old?
17.	What is the median age of the subjects in the sample?
18.	What is the standard deviation of the sample age values?
19.	What is the 70th percentile of the age values distribution?


```{r}
summary(wm)
summary(wm$ppage)
```


```{r}
sum(is.na(wm$ppage))
```


```{r}
wm <- wm %>%
    mutate(ppage2 = as.numeric(ppage))
```


```{r}
summary(wm$ppage2)
```


```{r}
wm$ppage[is.na(wm$ppage2)]
```


```{r}
wm <- wm %>%
    mutate(ppage = ifelse(ppage == "28 years", "28", ppage)) %>%
    mutate(ppage2 = as.numeric(ppage))
```


```{r}
summary(wm$ppage2)
```


```{r}
mean(wm$ppage2)
mean(wm$ppage2, na.rm = TRUE)
```


```{r}
quantile(wm$ppage2, na.rm = TRUE)
quantile(wm$ppage2, na.rm = TRUE, c(0.35,0.70))
```


```{r}
sd(wm$ppage2, na.rm = TRUE)
```


```{r}
wm %>%
    summarise(mean_age = mean(ppage2, na.rm=TRUE), 
    nobs = n(), 
    median = median(ppage2, na.rm=TRUE),
    percentiles = quantile(wm$ppage2, na.rm=TRUE, c(0.35,0.70)))
```


20.	What is mean age of the males in the sample?
21.	What is the median age of the males in the sample?
22.	   What is the 20th percentile of the age distribution for males?


```{r}
summary(wm$ppage2[wm$ppgender == "Male"])
quantile(wm$ppage2[wm$ppgender == "Male"], na.rm=TRUE, c(0.20,0.50))
```


```{r}
wm <- wm %>% select(-ppage)
names(wm)

wm <- wm %>% rename(ppage = ppage2)
names(wm)
```


```{r}
save(wm, file = "./OUTPUT/Assignment_1.RData")
```


```{r}
#savehistory("./OUTPUT/Assignment_1_words_matter.R")
```


# Section 2


```{r}
umaru <- read_csv("./DATA/umaru_data.csv")
```


23.	How many observations are in the dataset?
24.	How many variables are in the dataset?
25.	How many treatment groups are there?  (treat)
26.	What type of variable is age?
27.	How many observations are missing a value for age?
28.	What type of variable is time?


```{r}
umaru <- apply_labels(umaru,
                      age = "age in years",
                      time = "time to relapse/censoring in days") %>%
  mutate(relapse = factor(
    relapse,
    levels = c(0, 1),
    labels = c("Censored", "Relapsed")
  ))
```


```{r}
dim(umaru)
```


```{r}
unique(umaru$treat) %>% length()
```


```{r}
class(umaru$age)
```


```{r}
table(is.na(umaru$age))
```


```{r}
class(umaru$time)
```


```{r}
umaru <- umaru %>% mutate(treat = as.factor(treat))
```


```{r}
str(umaru$treat)
```


```{r}
umaru <- umaru %>%
    mutate(time = as.numeric(gsub(": days", " ", time)))
```


```{r}
#umaru <- umaru %>% mutate(time = parse_number(time))
```


29.	What proportion of the sample relapsed during the study 
   follow-up period?
30.	What was the average amount of time study participants were followed 
   until relapse or censoring?
31.	What is the median age of study participants?


```{r}
prop.table(table(umaru$relapse))
```


```{r}
mean(umaru$time, na.rm = TRUE)
```


```{r}
mean(umaru$age, na.rm = TRUE)
```


```{r}
umaru %>% 
    summarise(ir = sum(as.numeric(relapse == "Relapsed"))/sum(time, na.rm = T))
```


```{r}
umaru %>% 
    summarise(ir = sum(as.numeric(relapse == "Relapsed"))/sum(time, na.rm = T)*365*100)
```


```{r}
require(survival)
umaru <- umaru %>% mutate(ttr = Surv(time, relapse))
```


```{r}
head(umaru$ttr)
```


```{r}
kmumaru <- survfit(ttr ~ 1, data = umaru)
```


```{r}
summary(kmumaru)
```


```{r}
save(wm, umaru, file = "./OUTPUT/Assignment_1.RData")
```


```{r}
save(wm, umaru, file = "./OUTPUT/ssignment_1.RData")
```

